Place of death in NRS COVID-19 deaths: data processing for analysis, 2023 data

### 2023-02-13 covid_nrs_place_of_death_data_2023.Rmd
editing historic periods for 2023 data (& beyond - fianlly writing abstract enough code!)

---
title: "Place of death in NRS COVID-19 deaths: data processing for analysis, 2022 data"
author: "Jan Savinc"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
  html_document:
  code_folding: hide
toc: true
toc_float: true
editor_options: 
  chunk_output_type: console
---

@@

# TODO: delete below, it's for the previous releases that included 2021 data also
# cell_ranges_for_causes_of_death_data <- list(
#   "All" = "A6:V110",
#   "Care home" = "A113:V217",
#   "Home & other non-institution" = "A220:V324",
#   "Hospital" = "A327:V431",
#   "Other" = "A434:V538"
# )

cell_ranges_for_causes_of_death_data <- list(
  "All" = "A6:V58",
  "Care home" = "A61:V113",
  "Home & other non-institution" = "A116:V168",
  "Hospital" = "A171:V223",
  "Other" = "A226:V278"
)

weekly_deaths_cause_of_death_2022_onwards <-
  map_dfr(
    .x = cell_ranges_for_causes_of_death_data,
    .f = ~readxl::read_excel(files_weekly_deaths$main_report, sheet = sheet_cause_of_death_data, range = .x), 
    .id = "place_of_death"
  ) %>%
  janitor::clean_names() %>%
  rename(year = registration_year, date_w_c = week_beginning) %>%
  select(-matches("_excess$|_average$")) %>%  # we can compute excess & averages later
  pivot_longer(cols = 5:ncol(.), names_to = "name", values_to = "number_of_deaths") %>%
  filter(!str_detect(name, "all_causes")) %>%  # we can calculate that from the others
  mutate(
    year = as.integer(year),
    cause_of_death = str_remove_all(name, "_five_year_average|_deaths"),
    cause_of_death = return_pretty_cause_of_death(cause_of_death),
    date_w_c = as.Date(date_w_c, origin = "1899-12-30"),
    week_number_run_over = compute_week_run_over(week_number, year, start_year = 2020)  # invent a "run-over" week number - weeks 54+ are in 2021, weeks 106+ are in 2022
  ) %>%
  select(-name) %>%
  filter(!is.na(number_of_deaths))  # remove blank entries - these are for the current year but for dates in the future
```

@@

@@

Historical here refers to the 2015-2019 period for 2020 & 2021, and 2016-2021 (skipping 2020) for 2022, useful for a comparison with the historical context of the pre-pandemic years. (There is another 'historical' dataset, which includes cause of death figures for 2020 and for 2021.)

@@

Because the data from July 2022 onwards are split up so that I load the 2020 and 2021 from different sources, I'll compute the

@@

```{r}
weekly_deaths_cause_of_death <-
  bind_rows(
    weekly_deaths_cause_of_death_2022_onwards,
    weekly_deaths_cause_of_death_2021,
    weekly_deaths_cause_of_death_2020 %>% 
      select(-deaths_mean) %>%
      filter(cause_of_death != "All")  # don't need all cause data in causes breadkown!
  )

@@

Note also that since 2022, the 5-year historical period has been defined as 2016-2021, with 2020 excluded, so we need to calculate two weekly averages!

```{r, warning=FALSE}
historical_weekly_average_overall <-
  bind_rows(
    ## 2015-2019 period
    weekly_deaths_2015_2019_overall %>%
    filter(year!="average") %>%  # remove averages
    group_by(year, week_number) %>%  # remove week 53 if it was 0 for all places  - this means there was no week 53!
    filter(!all(number_of_deaths==0 & week_number==53)) %>%
    ungroup %>%
    create_week_53_data_for_years_with_52_weeks %>%
    group_by(week_number,place_of_death) %>% 
    summarise(
      across(number_of_deaths, .fns = list(mean=mean,min=min,max=max)),
      .groups = "drop"
    ) %>% 
    rename_all(~str_replace_all(.x, pattern = "number_of_deaths", replacement = "deaths")) %>%
    mutate(historic_period = historic_period_depending_on_year(2020))
    ,
    ## 2016-2021, excluding 2020
    bind_rows(
      ## 2016-2019
      weekly_deaths_2015_2019_overall %>%
        mutate(year = as.integer(year)) %>%
        filter(year %in% 2016:2019),
      ## 2021
      weekly_deaths_overall %>% 
        rename(number_of_deaths = deaths_all_causes) %>%
        filter(year == 2021 & age == "all" & sex == "all")
      ) %>%
    group_by(year, week_number) %>%  # remove week 53 if it was 0 for all places  - this means there was no week 53!
    filter(!all(number_of_deaths==0 & week_number==53)) %>%
    ungroup %>%
    create_week_53_data_for_years_with_52_weeks %>%
    group_by(week_number,place_of_death) %>% 
    summarise(
      across(number_of_deaths, .fns = list(mean=mean,min=min,max=max)),
      .groups = "drop"
    ) %>% 
    rename_all(~str_replace_all(.x, pattern = "number_of_deaths", replacement = "deaths")) %>%
    mutate(historic_period = historic_period_depending_on_year(2022))
  )

@@

deaths_mean_min_max <- function(data_tbl) {
  data_tbl %>%
    summarise(
      deaths_mean = mean(number_of_deaths),
      deaths_min = min(number_of_deaths),
      deaths_max = max(number_of_deaths),
      .groups = "drop"
    )
}

make_2016_to_2019_and_2021_data <- function(historical_data, current_data) {
  bind_rows(
      historical_data %>% filter(year %in% 2016:2019),
      current_data %>% filter(year == 2021) %>% rename(number_of_deaths = deaths_all_causes)
    )
}

historical_weekly_average_hb <-
  bind_rows(
    ## 2015-2019
    weekly_deaths_2015_2019_hb %>%
    create_week_53_data_for_years_with_52_weeks() %>%
    group_by(
      week_number, place_of_death, ref_area
    ) %>%
    deaths_mean_min_max %>%
    mutate(historic_period = historic_period_depending_on_year(2020))
    ,
    ## 2016-2019, 2021
    make_2016_to_2019_and_2021_data(
      historical_data = weekly_deaths_2015_2019_hb, 
      current_data = weekly_deaths_hb
      ) %>%
    create_week_53_data_for_years_with_52_weeks() %>%
    group_by(
      week_number, place_of_death, ref_area
    ) %>%
    deaths_mean_min_max %>%
    mutate(historic_period = historic_period_depending_on_year(2022))
  )

historical_weekly_average_la <-
  bind_rows(
    ## 2015-2019
    weekly_deaths_2015_2019_la %>%
    create_week_53_data_for_years_with_52_weeks() %>%
    group_by(
      week_number, place_of_death, ref_area
    ) %>%
    deaths_mean_min_max %>%
    mutate(historic_period = historic_period_depending_on_year(2020))
    ,
    ## 2016-2019, 2021
    make_2016_to_2019_and_2021_data(
      historical_data = weekly_deaths_2015_2019_la, 
      current_data = weekly_deaths_la
      ) %>%
    create_week_53_data_for_years_with_52_weeks() %>%
    group_by(
      week_number, place_of_death, ref_area
    ) %>%
    deaths_mean_min_max %>%
    mutate(historic_period = historic_period_depending_on_year(2022))
  )

historical_weekly_average_sex_age <-
  bind_rows(
    ## 2015-2019
    weekly_deaths_2015_2019_sex_age %>%
    create_week_53_data_for_years_with_52_weeks() %>%
    group_by(
      week_number, place_of_death, age, sex
    ) %>%
    deaths_mean_min_max %>%
    mutate(historic_period = historic_period_depending_on_year(2020))
    ,
    ## 2016-2019, 2021
    make_2016_to_2019_and_2021_data(
      historical_data = weekly_deaths_2015_2019_sex_age, 
      current_data = weekly_deaths_sex_age
      ) %>%
    create_week_53_data_for_years_with_52_weeks() %>%
    group_by(
      week_number, place_of_death, age, sex
    ) %>%
    deaths_mean_min_max %>%
    mutate(historic_period = historic_period_depending_on_year(2022))
  )

historical_weekly_average_age <-
  bind_rows(
    ## 2015-2019
    weekly_deaths_2015_2019_age %>%
    create_week_53_data_for_years_with_52_weeks() %>%
    group_by(
      week_number, place_of_death, age
    ) %>%
    deaths_mean_min_max %>%
    mutate(historic_period = historic_period_depending_on_year(2020))
    ,
    ## 2016-2019, 2021
    make_2016_to_2019_and_2021_data(
      historical_data = weekly_deaths_2015_2019_age, 
      current_data = weekly_deaths_age
      ) %>%
    create_week_53_data_for_years_with_52_weeks() %>%
    group_by(
      week_number, place_of_death, age
    ) %>%
    deaths_mean_min_max %>%
    mutate(historic_period = historic_period_depending_on_year(2022))
  )

historical_weekly_average_sex <-
  bind_rows(
    ## 2015-2019
    weekly_deaths_2015_2019_sex %>%
    create_week_53_data_for_years_with_52_weeks() %>%
    group_by(
      week_number, place_of_death, sex
    ) %>%
    deaths_mean_min_max %>%
    mutate(historic_period = historic_period_depending_on_year(2020))
    ,
    ## 2016-2019, 2021
    make_2016_to_2019_and_2021_data(
      historical_data = weekly_deaths_2015_2019_sex, 
      current_data = weekly_deaths_sex
      ) %>%
    create_week_53_data_for_years_with_52_weeks() %>%
    group_by(
      week_number, place_of_death, sex
    ) %>%
    deaths_mean_min_max %>%
    mutate(historic_period = historic_period_depending_on_year(2022))
  )

# TODO: include the split between historical periods here also
compute_historical_annual_average_deaths <- function(data_tbl, ...) {
  grouping_vars <- enquos(...)
  data_tbl %>%
    group_by(year, place_of_death, historic_period, !!!grouping_vars) %>%
    summarise(  # calculate annual totals first...
    number_of_deaths = sum(number_of_deaths), 
    .groups = "keep"
  ) %>%
  ungroup(year) %>%  # remove year as grouping to average over year now
  summarise(
    annual_deaths_mean = mean(number_of_deaths),
    annual_deaths_min = min(number_of_deaths),
    annual_deaths_max = max(number_of_deaths),
    .groups = "drop"
  )
}

>>>>

---
title: "Place of death in NRS COVID-19 deaths: data processing for analysis, 2023 data"
author: "Jan Savinc"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
  html_document:
  code_folding: hide
toc: true
toc_float: true
editor_options: 
  chunk_output_type: console
---

@@

## updated for 2023 which contains both 2022 and 2023
cell_ranges_for_causes_of_death_data <- list(
  "All" = "A6:V110",
  "Care home" = "A113:V217",
  "Home & other non-institution" = "A220:V324",
  "Hospital" = "A327:V431",
  "Other" = "A434:V538"
)

## this was for only 2022 data
# cell_ranges_for_causes_of_death_data <- list(
#   "All" = "A6:V58",
#   "Care home" = "A61:V113",
#   "Home & other non-institution" = "A116:V168",
#   "Hospital" = "A171:V223",
#   "Other" = "A226:V278"
# )

# TODO: this will eventually only have 2023 data, change when that happens
weekly_deaths_cause_of_death_2022_onwards <-
  map_dfr(
    .x = cell_ranges_for_causes_of_death_data,
    .f = ~readxl::read_excel(files_weekly_deaths$main_report, sheet = sheet_cause_of_death_data, range = .x), 
    .id = "place_of_death"
  ) %>%
  janitor::clean_names() %>%
  rename(year = registration_year, date_w_c = week_beginning) %>%
  select(-matches("_excess$|_average$")) %>%  # we can compute excess & averages later
  pivot_longer(cols = 5:ncol(.), names_to = "name", values_to = "number_of_deaths") %>%
  filter(!str_detect(name, "all_causes")) %>%  # we can calculate that from the others
  mutate(
    year = as.integer(year),
    cause_of_death = str_remove_all(name, "_five_year_average|_deaths"),
    cause_of_death = return_pretty_cause_of_death(cause_of_death),
    date_w_c = as.Date(date_w_c, origin = "1899-12-30"),
    week_number_run_over = compute_week_run_over(week_number, year, start_year = 2020)  # invent a "run-over" week number - weeks 54+ are in 2021, weeks 106+ are in 2022
  ) %>%
  select(-name) %>%
  filter(!is.na(number_of_deaths))  # remove blank entries - these are for the current year but for dates in the future
```

@@

#### 2022

```{r}
# TODO: deal with 2022 data here, now that it's 'historical'
cell_ranges_for_causes_of_death_data_2022 <- list(
  "All" = "A6:V58",
  "Care home" = "A61:V113",
  "Home & other non-institution" = "A116:V168",
  "Hospital" = "A171:V223",
  "Other" = "A226:V278"
)

```

@@

From the [methodology for computing excess deaths during the covid pandemic](https://www.nrscotland.gov.uk/files//statistics/covid19/covid-deaths-methodology-excess-deaths.pdf):

> For 2022, the excess deaths calculation would usually incorporate the mortality data for the most recent 5 year period (2017 – 2021). However, given the unusual nature of the 2020 data and the decision previously taken for 2021, it has been agreed with ONS that the excess deaths calculation will compare 2022 mortality data with the average of 2016, 2017, 2018, 2019 and 2021, and therefore continue to exclude 2020.
Similarly for 2023, the five year average will again leave out 2020 and will be built from the data from 2017, 2018, 2019, 2021 and 2022.

'Historical' here refers to the 5-year period preceding each year, useful for a historic comparison. The historic period depedning on year has been defined as:

| year|historic_period      |
|----:|:--------------------|
| 2020|2015-2019            |
| 2021|2015-2019            |
| 2022|2016-2019, 2021      |
| 2023|2017-2019, 2021-2022 |

There is a second sense of 'historical' here, which denotes cause of death figures for 2020 to 2022; these were only provided as 5-year averages for 2015-2019, whereas we have a per-cause breakdown from 2020 onwards - to combine these, a 5-year historical average needs to be recomputed using a combination of the 2015-2019 data and the 2020+ data (using weighted averages).

@@

Because the data from July 2022 onwards are split up so that I load the 2020 and 2021 from different sources, I'll compute the **Note: I don't remember what came after this...**

@@

```{r}
weekly_deaths_cause_of_death <-
  bind_rows(
    weekly_deaths_cause_of_death_2022_onwards,
    weekly_deaths_cause_of_death_2021,
    weekly_deaths_cause_of_death_2020 %>% 
      select(-deaths_mean) %>%
      filter(cause_of_death != "All")  # don't need all cause data in causes breakdown!
  )

@@

Since 2021, the 5-year historic context has been amended to exclude 2020 (see [above section on historical cause of death data](#historical-causes-of-death)). Because of this, the weekly average needs to be recalculated depending on which historic period is of interest.

## Historic periods by year

This is a reference table for the year of data, its historic period, and a nested list of vectors of years of historical data.

```{r}
historic_period_by_year <-
  tibble(year = 2020:latest_year) %>%
  mutate(
    historic_period = historic_period_depending_on_year(year),
    years_in_historic_period = map(.x = as.integer(str_sub(historic_period, 1, 4)), .f = generate_5_years_historic_period_excl_2020)
    )
```

## Historic averages, overall

```{r, warning=FALSE}
## helper functions to filter historic data for different 
get_historic_and_current_data_from_year_range <- function(historic_data, current_data, year_range) {
  year_range <- year_range[year_range!=2020]  # remove 2020 from range
  historic_range <- year_range[year_range < 2020]
  current_range <- year_range[year_range >= 2020]
  bind_rows(
    historic_data %>% 
      filter(year %in% historic_range),
    current_data %>% 
      filter(year %in% current_range) %>% 
      rename(number_of_deaths = deaths_all_causes)  # also rename to historic data format
  )
}

historical_weekly_average_overall <-
  pmap_dfr(
    .l = historic_period_by_year,
    .f = function(year, historic_period, years_in_historic_period) {
      get_historic_and_current_data_from_year_range(
        historic_data = 
          weekly_deaths_2015_2019_overall %>%
          filter(year!="average") %>%
          mutate(year = as.integer(year))
          ,
        current_data = 
          weekly_deaths_overall %>%
          filter(age == "all" & sex == "all"),
        year_range = years_in_historic_period
      ) %>%
      group_by(year, week_number) %>%  # remove week 53 if it was 0 for all places  - this means there was no week 53!
      filter(!all(number_of_deaths==0 & week_number==53)) %>%
      ungroup %>%
      create_week_53_data_for_years_with_52_weeks %>%
      group_by(week_number,place_of_death) %>% 
      summarise(
        across(number_of_deaths, .fns = list(mean=mean,min=min,max=max)),
        .groups = "drop"
      ) %>% 
      rename_all(~str_replace_all(.x, pattern = "number_of_deaths", replacement = "deaths")) %>%
      mutate(historic_period = historic_period)
    }
  ) %>%
  distinct()

@@

deaths_mean_min_max <- function(data_tbl) {
  data_tbl %>%
    summarise(
      deaths_mean = mean(number_of_deaths),
      deaths_min = min(number_of_deaths),
      deaths_max = max(number_of_deaths),
      .groups = "drop"
    )
}
```

## Historic averages, HB, LA, sex, age

```{r}
historical_weekly_average_hb <-
  pmap_dfr(
    .l = historic_period_by_year,
    .f = function(year, historic_period, years_in_historic_period) {
      get_historic_and_current_data_from_year_range(
        historic_data = weekly_deaths_2015_2019_hb, 
        current_data = weekly_deaths_hb,
        year_range = years_in_historic_period
      ) %>%
      create_week_53_data_for_years_with_52_weeks() %>%
      group_by(
        week_number, place_of_death, ref_area
      ) %>%
      deaths_mean_min_max %>%
      mutate(historic_period = historic_period)
    }
  ) %>% 
  distinct  # the 2020 and 2021 data get identical historical periods & values, so I remove duplicates

historical_weekly_average_la <-
  pmap_dfr(
    .l = historic_period_by_year,
    .f = function(year, historic_period, years_in_historic_period) {
      get_historic_and_current_data_from_year_range(
        historic_data = weekly_deaths_2015_2019_la, 
        current_data = weekly_deaths_la,
        year_range = years_in_historic_period
      ) %>%
      create_week_53_data_for_years_with_52_weeks() %>%
      group_by(
        week_number, place_of_death, ref_area
      ) %>%
      deaths_mean_min_max %>%
      mutate(historic_period = historic_period)
    }
  ) %>% 
  distinct  # the 2020 and 2021 data get identical historical periods & values, so I remove duplicates

historical_weekly_average_sex_age <-
  pmap_dfr(
    .l = historic_period_by_year,
    .f = function(year, historic_period, years_in_historic_period) {
      get_historic_and_current_data_from_year_range(
        historic_data = weekly_deaths_2015_2019_sex_age, 
        current_data = weekly_deaths_sex_age,
        year_range = years_in_historic_period
      ) %>%
      create_week_53_data_for_years_with_52_weeks() %>%
      group_by(
        week_number, place_of_death, age, sex
      ) %>%
      deaths_mean_min_max %>%
      mutate(historic_period = historic_period)
    }
  ) %>% 
  distinct  # the 2020 and 2021 data get identical historical periods & values, so I remove duplicates

historical_weekly_average_age <-
  pmap_dfr(
    .l = historic_period_by_year,
    .f = function(year, historic_period, years_in_historic_period) {
      get_historic_and_current_data_from_year_range(
        historic_data = weekly_deaths_2015_2019_age, 
        current_data = weekly_deaths_age,
        year_range = years_in_historic_period
      ) %>%
      create_week_53_data_for_years_with_52_weeks() %>%
      group_by(
        week_number, place_of_death, age
      ) %>%
      deaths_mean_min_max %>%
      mutate(historic_period = historic_period)
    }
  ) %>% 
  distinct  # the 2020 and 2021 data get identical historical periods & values, so I remove duplicates

historical_weekly_average_sex <-
  pmap_dfr(
    .l = historic_period_by_year,
    .f = function(year, historic_period, years_in_historic_period) {
      get_historic_and_current_data_from_year_range(
        historic_data = weekly_deaths_2015_2019_sex, 
        current_data = weekly_deaths_sex,
        year_range = years_in_historic_period
      ) %>%
      create_week_53_data_for_years_with_52_weeks() %>%
      group_by(
        week_number, place_of_death, sex
      ) %>%
      deaths_mean_min_max %>%
      mutate(historic_period = historic_period)
    }
  ) %>% 
  distinct  # the 2020 and 2021 data get identical historical periods & values, so I remove duplicates
```

## Historic averages, annual average

```{r}
# TODO: include the split between historical periods here also
compute_historical_annual_average_deaths <- function(data_tbl, ...) {
  grouping_vars <- enquos(...)
  data_tbl %>%
    group_by(year, place_of_death, historic_period, !!!grouping_vars) %>%
    summarise(  # calculate annual totals first...
    number_of_deaths = sum(number_of_deaths), 
    .groups = "keep"
  ) %>%
  ungroup(year) %>%  # remove year as grouping to average over year now
  summarise(
    annual_deaths_mean = mean(number_of_deaths),
    annual_deaths_min = min(number_of_deaths),
    annual_deaths_max = max(number_of_deaths),
    .groups = "drop"
  )
}
